Multilevel computation of glacier geometry \\ from Stokes dynamics

### 2021-12-26 paper/msg.tex
Edits.

\begin{figure}[t]
\begin{center}
\includegraphics[width=0.7\textwidth]{genfigs/stokesdomain.pdf}
\end{center}
\caption{In the SIGP the CMB $a$ (arrows; down for accumulation) and bed elevation $b$ (solid) are given data on a map-plane region $\Omega \subset \RR^d$.  The solution includes a surface elevation $s$ (dashed) on $\Omega$, such that $s\ge b$, plus the ice velocity $\bu$ and pressure $p$ on the icy domain $\Lambda_s = \{b < z < s\} \subset \RR^{d+1}$.}
\label{fig:stokesdomain}
\end{figure}

@@

In computations we will use constants $\nn=3$, ice density $\rhoi=910 \,\text{kg}\,\text{m}^{-3}$ \cite{Huybrechtsetal1996}, and gravity $\bg=\left<0,0,-g\right>$, with $g=9.81\,\text{m}\,\text{s}^{-2}$.  The ice hardness $B_\nn$ is a constant because we assume isothermal conditions \cite{GreveBlatter2009}, with the value $B_3=6.8082\times 10^7\,\text{Pa}\,\text{s}^{1/3}$ \cite{Huybrechtsetal1996} used in computations.

For the $\nn=1$ (Newtonian) Stokes equations one would write \eqref{eq:viscflowlaw} as $\tau = 2\nu D\bu$ with viscosity $\nu>0$, but exponents $\nn>1$ imply an effective viscosity function of $|D\bu|$.  This would be singular in the limit of small strain rates, and so, motivated by the actual finite viscosity of glacier ice \cite{GreveBlatter2009}, we define the regularized effective viscosity using $\eps>0$,
\begin{equation}
\nu_\eps(|D\bu|) = \frac{1}{2} B_\nn \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}, \label{eq:regeffvisc}
\end{equation}
where $\pp=(1/\nn)+1$, with $\pp=4/3$ in computations.  The constant $D_0$ defines a strain-rate scale for glacier flow; values $D_0 = 1 \,\text{a}^{-1}$ and $\eps = 10^{-4}$ are used in computations.

Our SIGP model uses dynamic boundary conditions for isolated, grounded, and non-sliding glaciers.  (Floating and sliding cases are topics for additional research.)  In addition to the already-stated assumption that the top and bottom boundaries of $\Lambda_s$ can be identified, we further assume these surfaces have well-defined tangents.  On the base we impose no slip:
\begin{equation}
\bu = \bzero  \qquad\qquad \text{\emph{base} } \Gamma_0. \label{eq:basebc}
\end{equation}
On the other sub-aerial surfaces we set a condition of zero applied stress,
\begin{equation}
\left(2 \nu_\eps(|D\bu|) D\bu - pI\right) \bn = \bzero  \qquad \qquad \text{\emph{top and cliffs (if present)}} \label{eq:topbc}
\end{equation}
where $\bn$ is any normal to $\overline{\partial} \Lambda$.  (A nonzero atmospheric pressure at the surface is straightforward to apply if desired.)  Note that the ice flow extends in the horizontal direction until a free boundary at the glacier margin is reached, but that $\grad s$ may become singular there.

The simultaneous determination of $\Lambda_s$ and $(\bu,p)$ is the goal of the SIGP model.  However, the above equations make no reference to the climate input function $a$, so we need another ``equation'', one which is actually an inequality.  Noting that the surface elevation is already defined on all of $\Omega$, with $s=b$ off the ice, we set
\begin{equation}
\bn_s = \left<-s_x,-s_y,1\right> \label{eq:surfacenormal}
\end{equation}
as an (un-normalized) top-surface normal almost everywhere on $\Omega$.  Furthermore we extend the surface value of the velocity to all of $\Omega$:
\begin{equation}
\bus(x,y) = \begin{cases} \bu(x,y,s(x,y)), & s(x,y) > b(x,y), \\
                     \bzero, & \text{elsewhere}. \end{cases} \label{eq:surfacevelocity}
\end{equation}
A discontinuity of this function is allowed at the ice margin.

The steady-state surface kinematical equation (SKE) \cite[equation (5.21)]{GreveBlatter2009} is the needed additional aspect of mass conservation:
\begin{equation}
\bus \cdot \bn_s + a = 0 \qquad \text{\emph{on the ice}}. \label{eq:ske}
\end{equation}
Note that $a(x,y)$ is the \emph{vertical} ice thickness added per time, while some references use a distinct quantity, namely the thickness added perpendicularly to the ice surface, significantly different if $\grad s$ is large.

Though kinematical balance \eqref{eq:ske} applies only on the ice, $\bus \cdot \bn_s + a \le 0$ applies everywhere in $\Omega$ because $a$ is nonpositive in steady state in ice-free locations.  In fact, when combined with the constraint that $s\ge b$ everywhere on $\Omega$, SKE \eqref{eq:ske} is part of an infinite-dimensional nonlinear complementarity problem (NCP) \cite{Bueler2021conservation}.  An NCP on a finite-dimensional vector space $V=\RR^k$ combines the three statements
\begin{equation}
x\ge 0, \quad f(x)\ge 0, \quad x f(x)=0 \label{eq:ncp}
\end{equation}
for all $x\in V$, where $f:V\to V$ \cite{FacchineiPang2003}.

@@

The solution of \eqref{eq:strongform} is a triple of functions $s(x,y)$, $\bu(x,y,z)$, $p(x,y,z)$ on $\Omega,\Lambda_s,\Lambda_s$, respectively.  However, as the domain on which $\bu,p$ are defined is only known via the solution, \eqref{eq:strongform} is at best an incomplete description.  The weak form in the next section, using a solution operator which maps $s$ to $\bu|_s$, will partially address this concern.

However, whether in strong or weak form, inequality-constrained system \eqref{eq:strongform} has a largely-unknown theory regarding well-posedness and solution regularity.  Certain well-posedness theory is known for the SIA version of this problem, with existence established by \cite{JouvetBueler2012}, including uniqueness in the flat bed case.

@@

After stating the SIGP weak form next in Section \ref{sec:weakido}, in Sections \ref{sec:fe}--\ref{sec:results} we will construct and demonstrate a multilevel scheme for the problem.  As noted in Section \ref{sec:intro}, numerical solutions have traditionally applied explicit time-stepping, even when computing steady states, splitting the dynamics and the SKE into sub-steps and using truncation to address the NCP \cite[for example]{Jouvetetal2008,Lengetal2012}.  By contrast, the implicit time-stepping model in Section \ref{sec:evolution} avoids time-splitting and is unconditionally stable.

@@

The weak form of the fixed-geometry Glen-Stokes model for ice flow, which extends the linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here.  Then we address the weak form of the SIGP.

Denote by $W^{k,r}$ the Sobolev space \cite{Evans2010} of functions with $k$ weak derivatives which are $r$th-power integrable.  Suppose for the moment that $\Lambda \subset \RR^{d+1}$ is a fixed, bounded domain on which we prescribe a Dirchlet condition $\bu=0$ on some positive-measure portion of $\partial\Lambda$, and otherwise suppose the boundary is stress-free (Neumann boundary).  Let $\pp=(1/\nn)+1$ as in \eqref{eq:regeffvisc}, and $\qq=(1-\pp^{-1})^{-1}=\nn+1$ the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $\nn=3$.  Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions, zero along the Dirichlet boundary, and let
\begin{equation}
\mathcal{M}_\Lambda = W_0^{1,\pp}(\Lambda)^{d+1} \times L^\qq(\Lambda)  \label{eq:mixed}
\end{equation}
be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes solution $(\bu,p) \in \mathcal{M}_\Lambda$ satisfies the weak form
\begin{equation}
F_\Lambda(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps(|D\bu|) D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
\end{equation}
for all $(\bv,q) \in \mathcal{M}_\Lambda$.  If $\bu,p$ are sufficiently regular then they satisfy strong equations \eqref{eq:forcebalance}--\eqref{eq:topbc}.

Jouvet and Rappaz \cite{JouvetRappaz2011} prove that this fixed-domain Glen-Stokes formulation is well-posed under the above assumptions if also the Neumann boundary of $\Lambda$ is $C^1$.  They show \eqref{eq:glenstokesweak} is equivalent to minimization of a convex and coercive functional over the divergence-free subspace, and that there is a unique pressure $p$.

However, we must go beyond the problem on predetermined ice geometry to solve the weak form of SIGP \eqref{eq:strongform}.  Again suppose $\Omega \subset \RR^d$ is a fixed map-plane domain with $C^1$ boundary, and that the bed elevation $b$ is in $W^{1,\qq}(\Omega)$.  The Sobolev space $W^{1,\qq}(\Omega)$ supports a linear trace operator which defines the symbol $|_{\partial \Omega}$ \cite[Section 5.5]{Evans2010}.  Let
\begin{equation}
\mathcal{K} = \{s \in W^{1,\qq}(\Omega) \,:\, s \ge b \, \text{ and } \, s\big|_{\partial\Omega} = b\big|_{\partial\Omega}\}  \label{eq:Kconstraintset}
\end{equation}
be the closed and convex set of admissible surface elevations of an isolated glacier or ice sheet.  From now on we also assume that $\nn>1$.  Because $\qq = \nn+1 > d$, each $s \in W^{1,\qq}(\Omega)$ is continuous \cite[Morrey's inequality, section 5.6.2]{Evans2010}, and so there are no cliffs in the $s$ geometry.  (Though it is not yet clear that the correct Sobolev space is identified in \eqref{eq:Kconstraintset}, existence for the simpler SIA model finds a power of the ice thickness, $u=(s-b)^{2q/(q-1)}$, in $\mathcal{K} = \{v \ge 0\} \subset W^{1,\qq}(\Omega)$ \cite{JouvetBueler2012}.)  Recalling definition \eqref{eq:lambdas}, we further suppose that
\begin{equation}
s\in \mathcal{K} \text{ defines an icy domain } \Lambda_s \text{ which has a $C^1$ upper surface.} \label{eq:quixotic}
\end{equation}
Assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, something we cannot offer.  However, in the SIA model the surface $s$ solves an elliptic equation, and $s$ is $C^1$ on $\{s>b\} \subset \Omega$ if $b$ is continuous \cite{JouvetBueler2012}.

We now define an ice dynamics operator (IDO), a map $\Phi$ which uses the trace of the velocity solution to the Glen-Stokes problem \eqref{eq:glenstokesweak}; see Figure \ref{fig:idoaction}.  For each $s \in \mathcal{K}$ the output $\Phi(s)$ is the negative normal component of the (unique) velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the upper surface of $\Lambda_s$, but extended by zero to all of $\Omega$:
\begin{equation}
\Phi(s) = \begin{cases} - \bu|_s \cdot \bn_s, & s > b, \\
                        0, & \text{otherwise}. \end{cases} \label{eq:ido}
\end{equation}
Here $\bu|_s$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.  Note that $\Phi(s)$ is a scalar function on $\Omega$ which may jump or be singular at the glacier margin even if $s$ is smooth on $\{s>b\}$.  (We will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$.)

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{genfigs/idoaction.pdf}
\end{center}
\caption{The IDO $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=- \bu|_s \cdot \bn_s$.  Both $s$ and $\Phi(s)$ are defined on all of $\Omega$, but $\Phi(s)$ may be discontinuous as shown.}
\label{fig:idoaction}
\end{figure}

We believe that formula \eqref{eq:ido} defines a nonlinear operator $\Phi:\mathcal{K} \subset W^{1,\qq}(\Omega) \to (W^{1,\qq}(\Omega))^* = W^{-1,\pp}(\Omega)$.  This operator is \emph{expensive} and \emph{nonlocal}.  Evaluating $\Phi(s)$ requires solving Glen-Stokes problem \eqref{eq:glenstokesweak} on $\Lambda_s \subset \RR^{d+1}$ defined in \eqref{eq:lambdas}, and furthermore the solution $\bu$ is influenced everywhere by changes in the icy domain geometry.  In particular, if $\tilde s=s + r$ defines $\Lambda_{\tilde s} \supset \Lambda_s$, for $r\in C_c^\infty(\Omega)$ with $r\ge 0$ supported inside $\{s>b\}$, then the Stokes solutions are generically different \emph{everywhere}: $\tilde\bu - \bu \ne 0$ on all of $\Lambda_s$, including the trace on the upper surface.  It follows that an FE discretization (Section \ref{sec:fe}) of $\Phi(s)$ will have a dense Jacobian matrix

By contrast, in the simpler isothermal, nonsliding SIA model the operator is a nonlinear, second-order differential operator,
\begin{equation}
\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right], \label{eq:phisia}
\end{equation}
for $\gamma = 2(\rhoi g)^{\nn} (B_\nn)^{-\nn} > 0$ constant and $\grad_2 = (\partial_x,\partial_y)$.  Thus an FE discretization of $\Phi_{\text{SIA}}(s)$ has sparse Jacobian.  However, $\Phi_{\text{SIA}} \approx \Phi$ for shallow glaciers to the degree that the lubrication approximation \cite{Acheson1990} is valid, and in this sense we expect the Jacobian of $\Phi$ to be well-approximated by a banded matrix.

@@

For now we re-write \eqref{eq:strongform} as a clean NCP with a hidden dynamical problem:
\begin{align}
s - b &\ge 0  \label{eq:idostrongform} \\
\Phi(s) - a &\ge 0 \notag \\
(s - b) (\Phi(s) - a) &= 0 \notag
\end{align}
Here the statements hold on all of $\Omega$ and the roles of the data $a,b$ are evident.  The solution is the ice surface $z=s(x,y)$ while the variables $\bu,p$ are hidden within the evaluation of $\Phi(s)$.

To show the SIGP weak form we define a functional which is simply the dual of $\Phi$:
\begin{equation}
F(s)[r] = \ip{\Phi(s)}{r} = \int_\Omega \Phi(s)\, r \,dx dy. \label{eq:sigpfunctional}
\end{equation}
One computes $F(s)[r]$ by defining $\Lambda_s$ as in \eqref{eq:lambdas}, solving \eqref{eq:glenstokesweak}, evaluating the surface trace of the velocity to compute $\Phi(s)$, and then integrating the result against (i.e.~dual pairing with) the test function $r$.  The weak form itself is the following variational inequality (VI) \cite{KinderlehrerStampacchia1980} for $s\in\mathcal{K}$:
\begin{equation}
F(s)[r - s] \ge \ip{a}{r-s} \quad \text{for all $r \in \mathcal{K}$.}  \label{eq:sigpweakform}
\end{equation}
One can prove that if $s$ is $C^1$ and solves \eqref{eq:sigpweakform} then it solves strong forms \eqref{eq:strongform} and \eqref{eq:idostrongform} \cite{Bueler2021conservation}.  In particular the strong SKE \eqref{eq:ske} holds on $s>b$; it is the interior condition of VI \eqref{eq:sigpweakform} \cite{KinderlehrerStampacchia1980}.

Speaking geometrically, \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a point where $\Phi(s)-a$ points directly into $\mathcal{K}$.  (The ``angle'' between $\Phi(s)-a$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.)  It is well-known that certain VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge \eqref{eq:sigpweakform} does \emph{not} arise in this way.  Existence has been proven for the SIA analog of VI \eqref{eq:sigpweakform} \cite{JouvetBueler2012}, but, as noted above, in that case the residual is locally-computable.  This model is also not a minimization, except in the flat-bed case.

@@

In the next three sections we will construct an iterative, multilevel finite element solver for SIGP weak form \eqref{eq:sigpweakform}.  Substantial performance and stability concerns related to evaluating $\Phi'$ are addressed in Section \ref{sec:smoothers}.

@@

Assume $\Omega \subset \RR^d$ is polygonal with triangulation $\mathcal{T}$.  (If $d=1$ then $\mathcal{T}$ denotes an interval decomposition of $\Omega$.)  Based on low expected regularity at the ice margin, we will represent surface elevations $s\in \mathcal{K}$ using the $P_1$ finite element (FE) space
\begin{equation}
\mathcal{V}^h = \{s \in C^0(\Omega) : s|_T \text{ is linear if } T \in \mathcal{T}\} \subset W^{1,\qq}(\Omega).
\end{equation}
Note \cite{JouvetBueler2012} makes the same FE choice for the corresponding SIA problem, and that, because of low regularity at the free boundary, $P_1$ is a common choice even for the classical obstacle problem with smooth data \cite{GraeserKornhuber2009}.

>>>>

\begin{figure}[t]
\begin{center}
\includegraphics[width=0.7\textwidth]{genfigs/stokesdomain.pdf}
\end{center}
\caption{In the SIGP the CMB $a$ (arrows; down for accumulation) and bed elevation $b$ (solid) are given data on a map-plane region $\Omega \subset \RR^d$ for $d=1,2$.  The solution includes a surface elevation $s$ (dashed) on $\Omega$, such that $s\ge b$, plus the ice velocity $\bu$ and pressure $p$ on the icy domain $\Lambda_s = \{b < z < s\} \subset \RR^{d+1}$.}
\label{fig:stokesdomain}
\end{figure}

@@

In computations we will use the following constants: Glen exponent $\nn=3$, ice density $\rhoi=910 \,\text{kg}\,\text{m}^{-3}$ \cite{Huybrechtsetal1996}, and gravity $\bg=\left<0,0,-g\right>$, with $g=9.81\,\text{m}\,\text{s}^{-2}$.  The ice hardness $B_\nn$ is a constant because we assume isothermal conditions \cite{GreveBlatter2009}, with value $B_3=6.8082\times 10^7\,\text{Pa}\,\text{s}^{1/3}$ \cite{Huybrechtsetal1996}.

For the $\nn=1$ (Newtonian) Stokes equations one would write \eqref{eq:viscflowlaw} as $\tau = 2\nu D\bu$ with viscosity $\nu>0$, but exponents $\nn>1$ suggest an effective viscosity function of $|D\bu|$.  This would be singular in the limit of small strain rates, and so, motivated by the actual finite viscosity of glacier ice \cite{GreveBlatter2009}, we define the following regularized effective viscosity using $\eps>0$,
\begin{equation}
\nu_\eps(|D\bu|) = \frac{1}{2} B_\nn \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}, \label{eq:regeffvisc}
\end{equation}
where $\pp=(1/\nn)+1$, with $\pp=4/3$ in computations.  The constant $D_0$ defines a strain-rate scale for glacier flow; values $D_0 = 1 \,\text{a}^{-1}$ and $\eps = 10^{-4}$ are used in computations.

Our SIGP model uses dynamic boundary conditions for an isolated, grounded, and non-sliding glaciers.  (Floating and sliding cases are topics for additional research.)  In addition to the already-stated assumption that the top and bottom boundaries of $\Lambda_s$ can be identified, we further assume these surfaces have well-defined tangents.  On the base we impose no slip:
\begin{equation}
\bu = \bzero  \qquad\qquad \text{\emph{base} } \Gamma_0. \label{eq:basebc}
\end{equation}
On the sub-aerial surfaces we set a condition of zero applied stress,
\begin{equation}
\left(2 \nu_\eps(|D\bu|) D\bu - pI\right) \bn = \bzero  \qquad \qquad \text{\emph{upper ice surface}} \label{eq:topbc}
\end{equation}
where $\bn$ is any normal to $\partial \Lambda_s$.  (A nonzero atmospheric pressure at the surface is straightforward to apply if desired.)  Note that the ice flow extends in the horizontal direction until a free boundary at the glacier margin is reached.  Generally $\grad s$ is singular at a glacier margin.

The simultaneous determination of $\Lambda_s$ and $(\bu,p)$ is the goal of the SIGP model.  However, the above equations make no reference to the climate input function $a$, so we need another ``equation'', one which is fundamentally an inequality as well.  Noting that the surface elevation is already defined on all of $\Omega$, with $s=b$ off the ice, we set
\begin{equation}
\bn_s = \left<-s_x,-s_y,1\right> \label{eq:surfacenormal}
\end{equation}
as an (un-normalized) top-surface normal defined almost everywhere on $\Omega$.  Furthermore we extend the surface value of the velocity to all of $\Omega$:
\begin{equation}
\bus(x,y) = \begin{cases} \bu(x,y,s(x,y)), & s(x,y) > b(x,y), \\
                     \bzero, & \text{elsewhere}. \end{cases} \label{eq:surfacevelocity}
\end{equation}
A discontinuity of this function is (generally) expected at the ice margin.

The steady-state surface kinematical equation (SKE) \cite[equation (5.21)]{GreveBlatter2009} is the needed additional equation:
\begin{equation}
\bus \cdot \bn_s + a = 0 \qquad \text{\emph{on the ice}}. \label{eq:ske}
\end{equation}
Note that $a(x,y)$ is the \emph{vertical} ice thickness added per time \cite{GreveBlatter2009}.  Though kinematical balance \eqref{eq:ske} applies only on the ice, $\bus \cdot \bn_s + a \le 0$ applies everywhere in $\Omega$ because $a$ is nonpositive in steady state in ice-free locations.  In fact, when combined with the constraint that $s\ge b$ everywhere on $\Omega$, SKE \eqref{eq:ske} is part of an infinite-dimensional nonlinear complementarity problem (NCP) \cite{Bueler2021conservation}.  By definition, an NCP on a vector space $V$ combines the three statements
\begin{equation}
x\ge 0, \quad f(x)\ge 0, \quad x f(x)=0 \label{eq:ncp}
\end{equation}
for all $x\in V$, where $f:V\to V$ \cite{FacchineiPang2003}.

@@

The solution of \eqref{eq:strongform} is a triple of functions $s(x,y)$, $\bu(x,y,z)$, $p(x,y,z)$ on $\Omega,\Lambda_s,\Lambda_s$, respectively.  However, as the domain on which $\bu,p$ are defined is only known via the solution, \eqref{eq:strongform} is at best an incomplete description.  (The weak form in the next section, using a solution operator which maps $s$ to $\bu|_s$, will address this concern.)  However, whether in strong or weak form, inequality-constrained system \eqref{eq:strongform} has a largely-unresolved theory regarding well-posedness and solution regularity.  Certain well-posedness theory is known for the SIA version of this problem, with existence established by \cite{JouvetBueler2012}, including uniqueness in the flat bed case.

@@

After stating the SIGP weak form next in Section \ref{sec:weakido}, in Sections \ref{sec:fe}--\ref{sec:results} we will construct and demonstrate a multilevel scheme for the problem.  Noting that numerical solutions have traditionally applied explicit time-stepping, even when computing steady states, splitting the dynamics and the SKE into sub-steps and using truncation to address the NCP \cite[for example]{Jouvetetal2008,Lengetal2012}, the implicit time-stepping model in Section \ref{sec:evolution} avoids time-splitting and is unconditionally stable.

@@

The weak form of the fixed-geometry Glen-Stokes model for ice flow, which extends the linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here.  Then we address the weak form of the geometry-determining SIGP.

Denote by $W^{k,r}$ the Sobolev space \cite{Evans2010} of functions with $k$ weak derivatives which are $r$th-power integrable.  Suppose for the moment that $\Lambda \subset \RR^{d+1}$ is a fixed, bounded domain on which we prescribe a Dirchlet condition $\bu=0$ on some positive-measure portion of $\partial\Lambda$, and otherwise suppose the boundary is stress-free (Neumann boundary).  Let $\pp=(1/\nn)+1$ as in \eqref{eq:regeffvisc}, and $\qq=(1-\pp^{-1})^{-1}=\nn+1$ the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $\nn=3$.  Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions, zero along the Dirichlet boundary, and let
\begin{equation}
\mathcal{M}_\Lambda = W_0^{1,\pp}(\Lambda)^{d+1} \times L^\qq(\Lambda)  \label{eq:mixed}
\end{equation}
be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes solution $(\bu,p) \in \mathcal{M}_\Lambda$ satisfies the weak form
\begin{equation}
F_\Lambda(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps(|D\bu|) D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
\end{equation}
for all $(\bv,q) \in \mathcal{M}_\Lambda$.  If $\bu,p$ are sufficiently regular then they satisfy strong equations \eqref{eq:forcebalance}--\eqref{eq:topbc}.  Jouvet and Rappaz \cite{JouvetRappaz2011} prove that this fixed-domain Glen-Stokes formulation is well-posed under the above assumptions if also the Neumann boundary of $\Lambda$ is $C^1$.  They show \eqref{eq:glenstokesweak} is equivalent to minimization of a convex and coercive functional over the divergence-free subspace, and that there is a unique pressure $p$.

However, we must go beyond predetermined ice geometry to solve the weak form of SIGP \eqref{eq:strongform}.  Again suppose $\Omega \subset \RR^d$ is a fixed map-plane domain with $C^1$ boundary, and that the bed elevation $b$ is in $W^{1,\qq}(\Omega)$.  The Sobolev space $W^{1,\qq}(\Omega)$ supports a linear trace operator which defines the symbol $|_{\partial \Omega}$ \cite[Section 5.5]{Evans2010}.  Let
\begin{equation}
\mathcal{K} = \{s \in W^{1,\qq}(\Omega) \,:\, s \ge b \, \text{ and } \, s\big|_{\partial\Omega} = b\big|_{\partial\Omega}\}  \label{eq:Kconstraintset}
\end{equation}
be the closed and convex set of admissible surface elevations of an isolated glacier or ice sheet.  From now on we also assume that $\nn>1$.  Because $\qq = \nn+1 > d$, each $s \in W^{1,\qq}(\Omega)$ is continuous \cite[Morrey's inequality, section 5.6.2]{Evans2010}, and so there are no cliffs in the $s$ geometry.  (Though it is not yet clear that the correct Sobolev space is identified in \eqref{eq:Kconstraintset}, existence for the simpler SIA model finds that a power of the ice thickness, $u=(s-b)^{2q/(q-1)}$, is in $\mathcal{K} = \{v \ge 0\} \subset W^{1,\qq}(\Omega)$ \cite{JouvetBueler2012}.)  Recalling definition \eqref{eq:lambdas}, we further suppose that
\begin{equation}
s\in \mathcal{K} \text{ defines an icy domain } \Lambda_s \text{ which has a $C^1$ upper surface.} \label{eq:quixotic}
\end{equation}
Assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, something we cannot offer.  However, in the SIA model the surface $s$ solves an elliptic equation, and $s$ is $C^1$ on $\{s>b\} \subset \Omega$ if $b$ is continuous \cite{JouvetBueler2012}.

We now define an ice dynamics operator (IDO), a map $\Phi$ which uses the trace of the velocity solution to the Glen-Stokes problem \eqref{eq:glenstokesweak}; see Figure \ref{fig:idoaction}.  For each $s \in \mathcal{K}$ the output $\Phi(s)$ is the negative normal component of the (unique) velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the upper surface of $\Lambda_s$, but extended by zero to all of $\Omega$:
\begin{equation}
\Phi(s) = \begin{cases} - \bu|_s \cdot \bn_s, & s > b, \\
                        0, & \text{otherwise}. \end{cases} \label{eq:ido}
\end{equation}
Here $\bu|_s$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.  Note that $\Phi(s)$ is a scalar function on $\Omega$ which may jump or be singular at the glacier margin even if $s$ is smooth on $\{s>b\}$.  We will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$.

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{genfigs/idoaction.pdf}
\end{center}
\caption{The IDO $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=- \bu|_s \cdot \bn_s$.  Both $s$ and $\Phi(s)$ are defined on all of $\Omega$, but $\Phi(s)$ is generally discontinuous.}
\label{fig:idoaction}
\end{figure}

We believe that formula \eqref{eq:ido} defines a nonlinear operator $\Phi:\mathcal{K} \subset W^{1,\qq}(\Omega) \to (W^{1,\qq}(\Omega))^* = W^{-1,\pp}(\Omega)$.  This operator is \emph{expensive} and \emph{nonlocal}.  Evaluating $\Phi(s)$ requires solving Glen-Stokes problem \eqref{eq:glenstokesweak} on $\Lambda_s \subset \RR^{d+1}$ defined in \eqref{eq:lambdas}, and furthermore the solution $\bu$ is influenced everywhere by changes in the icy domain geometry.  In particular, if $\tilde s=s + r$ defines $\Lambda_{\tilde s} \supset \Lambda_s$, for $r\in C_c^\infty(\Omega)$ with $r\ge 0$ supported inside $\{s>b\}$, then the Stokes solutions are generically different \emph{everywhere}; generically $\tilde\bu - \bu \ne 0$ a.e.~on $\Lambda_s$.  It follows that an FE discretization (Section \ref{sec:fe}) of $\Phi(s)$ will have a dense Jacobian matrix.

By contrast, in the simpler isothermal, nonsliding, $d=2$ SIA model the operator is a nonlinear, second-order differential operator,
\begin{equation}
\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right], \label{eq:phisia}
\end{equation}
for $\gamma = 2(\rhoi g)^{\nn} (B_\nn)^{-\nn} > 0$ constant and $\grad_2 = (\partial_x,\partial_y)$.  Thus an FE discretization of $\Phi_{\text{SIA}}(s)$ has sparse Jacobian.  However, $\Phi_{\text{SIA}} \approx \Phi$ for shallow glaciers to the degree that the lubrication approximation \cite{Acheson1990} is valid, and in fact for ice sheets we expect the Jacobian of $\Phi$ to be well-approximated by a banded matrix.

@@

For now we re-write \eqref{eq:strongform} as a cleaner NCP with a hidden dynamical problem:
\begin{align}
s - b &\ge 0  \label{eq:idostrongform} \\
\Phi(s) - a &\ge 0 \notag \\
(s - b) (\Phi(s) - a) &= 0 \notag
\end{align}
These statements hold on all of $\Omega$, and the roles of the data $a,b$ are evident, and the solution is the ice surface $z=s(x,y)$, but the variables $\bu,p$ are hidden within the evaluation of $\Phi(s)$.

To reveal the SIGP weak form we define a functional which is simply the dual of $\Phi$:
\begin{equation}
F(s)[r] = \ip{\Phi(s)}{r} = \int_\Omega \Phi(s)\, r \,dx dy \label{eq:sigpfunctional}
\end{equation}
% in here one could integrate-by-parts on the vertical velocity using incompressibility:
% u|_s . n|_s = u|_s s_x + v|_s s_y + w|_s = u|_s s_x + v|_s s_y - \int_b^s (u_x + v_y) dz
% and then integrate by parts in plane
where $r \in W^{1,q}(\Omega)$.  One computes $F(s)[r]$ by defining $\Lambda_s$ as in \eqref{eq:lambdas}, solving \eqref{eq:glenstokesweak}, evaluating the surface trace of the velocity to compute $\Phi(s)$, and then integrating the result against (i.e.~dual pairing with) the test function $r$.  The weak form itself is the following variational inequality (VI) \cite{KinderlehrerStampacchia1980} for $s\in\mathcal{K}$:
\begin{equation}
F(s)[r - s] \ge \ip{a}{r-s} \quad \text{for all $r \in \mathcal{K}$.}  \label{eq:sigpweakform}
\end{equation}
One can prove that if $s$ is $C^1$ and solves \eqref{eq:sigpweakform} then it solves strong forms \eqref{eq:ske} and \eqref{eq:idostrongform} \cite{Bueler2021conservation}.  Note that the strong SKE \eqref{eq:ske}, which holds on $s>b$, is the interior condition of VI \eqref{eq:sigpweakform} \cite{KinderlehrerStampacchia1980}.

Speaking geometrically, \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a point where $\Phi(s)-a$ points directly into $\mathcal{K}$.  (The ``angle'' between $\Phi(s)-a$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.)  It is well-known that certain VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge \eqref{eq:sigpweakform} does \emph{not} arise in this way.  Existence has been proven for the SIA analog of VI \eqref{eq:sigpweakform} \cite{JouvetBueler2012}, and this SIA model is not a minimization except in the flat-bed case.

@@

In the next three sections we will construct an iterative, multilevel finite element solver for SIGP weak form \eqref{eq:sigpweakform}.  Performance and stability concerns related to evaluating $\Phi'$ are addressed in Section \ref{sec:smoothers}.

@@

Assume $d=2$ and that $\Omega \subset \RR^d$ is polygonal with triangulation $\mathcal{T}$.  (If $d=1$ then $\mathcal{T}$ would denote an interval decomposition of $\Omega$.)  Based on low expected regularity at the ice margin, we will represent surface elevations $s\in \mathcal{K}$ using the $P_1$ finite element (FE) space
\begin{equation}
\mathcal{V}^h = \{s \in C^0(\Omega) : s|_T \text{ is linear if } T \in \mathcal{T}\} \subset W^{1,\qq}(\Omega).
\end{equation}
Note \cite{JouvetBueler2012} makes the same FE choice for the corresponding SIA problem, and that, because of low regularity at the free boundary, $P_1$ is a common choice even for the classical obstacle problem with smooth data \cite{GraeserKornhuber2009}.
